Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-21 dictionary.tex
bringing up to date changes made on sage cloud

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes}% showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
Assuming only that $G$ is a smooth commutative group scheme over $\Fq$ with finitely generated component group, this paper gives a simple cohomological description of invisible quasicharacter sheaves on $G$. 
%The most interesting case is that of etale commutative group schemes over $\Fq$.

@@

In order to describe the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there exists an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

@@

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to etale, commutative group schemes over $\Fq$.
First, consider the etale group scheme $\ZZ$ over $\Fq$.
While the $\ell$-adic character $\ZZ \to \EEx$ defined by
$1 \mapsto -1$, has finite order, the character defined by
$1 \mapsto 1+\ell$ does not, although its image is bounded;
on the other hand, the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$ has unbounded image.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
Second, consider the etale group scheme $\ZZ \times \ZZ$.
\todo{Build an example of an invisible quasicharacter sheaf here.}
Finally, consider the etale group scheme $\ZZ \times \ZZ$ over $\Fq$ with Galois action determined by $\sigma(n,m) = (m,n)$ when the restriction of $\sigma$ to $\mathbb{F}_{q^2}$ is non-trivial.
Although there are infinitly-many geometric points on this group, it has exactly one $\Fq$-rational point.
In this case, the group carries only one isomorphism class of quasicharacter sheaves -- in particular, it has no invisible quasicharacter sheaves -- so the dictionary is perfect, albeit trivial, in this case.

@@

We also want to situate our terminology in a historical context.
Some people refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}.
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves. 
In particular, all quasicharacter sheaves on such groups are visible.

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who made many helpful observations and suggestions.
%
Finally, we gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our weeklong stay in May 2012.

@@

\begin{definition}\label{def:trFrob}
The function
\begin{align*}
\TrFrob{G}: \QCiso{G} &\to G(\Fq)^* \\
\qcs{L} &\mapsto t_{\qcs{L}}
\end{align*}
is called \emph{Trace of Frobenius}.
\end{definition}

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $s_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $s_\mathcal{L} = s_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$, and that $s_{\qcs{L}\otimes \qcs{L'}} = s_{\qcs{L}} \, s_\qcs{L'}$
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.
We claim that $s_\qcs{L}$ is fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
\begin{align*}
S_G: \QCiso{G} &\to \Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx))\\
\qcs{L} &\mapsto s_{\qcs{L}}.
\end{align*}

\begin{definition}\label{def:invisible}
Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We say $\qcs{L}$ is \emph{visible} if $S_G(\qcs{L}) \in H^0(\Weil{},H^2(G(\bFq),\EEx))$ is trivial; otherwise, we say $\qcs{L}$ is \emph{invisible}.
\end{definition}

\subsection{Group cohomology}
%If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves. We will return to that case in Section~\ref{sec:etale}.

While neither $T_{G}$ nor $S_{G}$ are injective, their direct sum is.

%In order to relate $\QCiso{G}$ to $G(k)^*$, we need a simple result relating duals, invariants and coinvariants.

\begin{proposition}
The group homomorphism
\begin{align*}
R_{G} : \QCiso{G} &\to  \Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx) \oplus \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx))\\
\qcs{L} &\mapsto (s_{\qcs{L}}, 
\end{align*}
be the map formed from  . . .
Then $R_{G}$ is injective.
\end{proposition}

@@

\begin{lemma}\label{lem:etale-2cocycle}
    If $G$ is an \'etale commutative group scheme over $\Fq$
    then the group homomorphism
    \[
    S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)),
    \]
    defined in Section~\ref{sec:invisible}, is surjective.\todo{Kernel? Canonical section?}
\end{lemma}

\begin{proof}
Pick $[g]\in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$. Then $g\in Z^2(G(\bFq),\EEx)$ is fixed by Frobenius, so $\,^{\Frob{}}g = g h$ for some $h\in B^2(G(\bFq),\EEx)$. Accordingly, $h(x,y) = \frac{f(x+y)}{f(x) f(y)}$ for some $f : G(\bFq) \to \EEx$, so
\begin{equation}\label{gf}
g(\Frob{}(x),\Frob{}(y)) = g(x,y) \frac{f(x+y)}{f(x) f(y)}
\end{equation}
for all $x,y\in G(\bFq)$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$ and define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x \otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = g(x,y) 1 \otimes 1$.
Since $g\in Z^2(G(\bFq),\EEx)$, the indexed set of isomorphisms $\{ \mu_{x,y} \tq x,y \in G(\bFq) \}$ satisfies \ref{qc.2}.
Define $\phi_x : \gqcs{L}_{\Frob{}(x)}\to \gqcs{L}_x$ by $\phi_x(1) = f(x)$.
Then set of isomorphisms $\{ \phi_{x} \tq x \in G(\bFq) \}$ satisfies \ref{qc.3} by \eqref{gf}. In this way, the cocycle $g\in Z^0(\Weil{}, Z^2(G(\bFq),\EEx))$ defines a quasicharacter sheaf $\qcs{L}$. It is clear that $\qcs{L}$ maps to $[g] \in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$ under \eqref{2cocycle}.
\end{proof}

\begin{lemma}\label{lem:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then for every the $\chi \in G(\Fq)^*$ there is a visible quasicharacter sheaf $\qcs{L}$ on $G$ such that $\trFrob{\qcs{L}} = \chi$.
In particular, Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*,
\]
is surjective, even when restricted to visible quasicharacter sheaves.\todo{Kernel? Canonical section?}
\end{lemma}

@@

\begin{proposition} \label{prop:etale-iso}
	If $G$ is an \'etale commutative group scheme over $\Fq$
	and if $G(\bFq)$ is finitely generated,
  	then
  	\[
  	\QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx).
  	\]
  	Moreover, this isomorphism is functorial
  	in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives
\[
T_{G}: \QCiso{G}\to  \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx)),
\]
which is surjective, Lemma~\ref{lem:etale-trace}.
Combining this with
\[
S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
gives
\[
\QCiso{G} \mathop{\longrightarrow}\limits^{T_{G}\oplus S_{G}} \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))\oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
The Hochschild-Serre spectral sequence (for finitely generated abelian groups!) provides
\[
\Hh^2(G(\bFq)\rtimes \Weil{},\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
Composing these last two maps defines
\begin{equation}
\QCiso{G}\to \Hh^2(G(\bFq)\rtimes \Weil{},\EEx).
\end{equation}
To see that this homomorphism is an isomorphism,\todo{Clifton, you are here.}
\end{proof}

@@

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
%0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is exact.

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme over
  $\Fq$ then Trace of Frobenius
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is a functorial surjection with kernel $\Hh^0(\Weil{}, \Hh^2(\pi_0({\bar G}), \EE))$ and a canonical section.
\end{theorem}

@@

\end{document}
%sagemathcloud={"zoom_width":120}

>>>>

%%%%%%%%%%%%%%%%% PACKAGES %%%%%%%%%%%%%%%%%
% Encodings
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
% AMS packages
\usepackage{amsmath, amsthm, amssymb, todonotes, showkeys}
\usepackage[alphabetic]{amsrefs}
% Fonts
\usepackage{mathrsfs, yfonts}
\usepackage{geometry}
% TikZ
\usepackage{tikz}
\usetikzlibrary{shapes,arrows,calc,matrix}
\usepackage{tikz-cd}
% Hyperrefs
\usepackage{hyperref}

@@

\subsection*{Overview}
Deligne's function--sheaf dictionary translates characters of connected, commutative algebraic groups $G$ over finite fields $\Fq$ into certain local systems on G, from which the character of $G(\Fq)$ can be recovered using the action of Frobenius on $G$.
The local systems which appear in this dictionary are easily characterised and often referred to as \emph{character sheaves} on $G$.
The result is a perfect dictionary between characters of $G(\Fq)$ and character sheaves on G, with which the formidable artillery of algebraic geometry can be brought to bear on the study of characters of $G(\Fq)$.
Although this dictionary extends to the case when $G$ is not connected, and even to all smooth commutative group schemes over $\Fq$, the result is not perfect.
In this generality, the correct class of sheaves on G are called \emph{quasicharacter sheaves}; when $G$ is connected and of finite type, quasicharacter sheaves on $G$ specialize to character sheaves on $G$.
While every character of $G(\Fq)$ may be converted, canonically, into a quasicharacter sheaf on G from which the character of $G(\Fq)$ may be recovered using the trace of Frobenius on G, generally there exist non-trivial quasicharacter sheaves on G for which the trace of Frobenius is trivial; these are called \emph{invisible quasicharacter sheaves} on $G$.
Invisible quasicharacter sheaves on G are precisely the quasicharacter sheaves on $G$ that are lost in translation under the function-sheaf dictionary.
This paper gives a cohomological description of invisible quasicharacter sheaves on $G$
and a trick to make invisible quasicharacter sheaves disappear,
should the notion prove too troubling.

@@

In order to describe the main result of this paper with more precision, let us briefly review how the function--sheaf dictionary for connected commutative algebraic groups over finite fields \cite{deligne:SGA4.5}*{sommes trig.} translates characters into local systems.
Let $\Fq$ be a finite field and let $G$ be a connected commutative algebraic group over $\Fq$; fix an algebraic closure $\bFq$ of $\Fq$
and let $\bG$ be the base change of $G$ to $\bFq$.
Fix a prime $\ell$ invertible in $\Fq$ and let $\chi : G(\Fq) \to \EEx$ be any character.
Using little more than the Lang map for $G$, $\chi$ determines an $\ell$-adic local system $\qcs{L}$ on $G$ from which the character $\chi$ can be recovered using the Frobenius automorphism of $\bG$.
While this recovery process applies to all $\ell$-adic local systems $\qcs{L}$ on $G$, those that produce characters of $G(\Fq)$ are distinguished by a simple property: there \emph{exists} an isomorphism
\begin{equation}\label{introbox}
m^* \qcs{L} \iso \qcs{L} \boxtimes \qcs{L}
\end{equation}
 of $\ell$-adic local systems on $G$, where $m : G\times G \to G$ is the multiplication map for $G$. The function--sheaf dictionary for $G$ identifies a monoidal category of $\ell$-adic local systems on $G$ such that the group of isomorphism classes in this category is canonically identified with the $\ell$-adic character group of $G(\Fq)$, thus providing the geometrization and categorification of characters of $G(\Fq)$.

@@

Three examples will illustrate some of the issues that arise when extending the function--sheaf dictionary to etale, commutative group schemes over $\Fq$.
As we show in the paper, $\ell$-adic local systems on $G$ can only be used to geometrize characters of finite order; quasicharacter sheaves, however, provide the geometrization of all characters of $G(\Fq)$.
In particular, we will see how to uniquely geometrize the character $\ZZ \to \EEx$ defined by
$1 \mapsto \ell$, which has unbounded image,
using quasicharacter sheaves on the etale group scheme $\ZZ$ with trivial Galois action.
All quasicharacter sheaves on this group are visible.
Likewise, the dictionary is perfect for the
etale group scheme $\ZZ$ with Galois action determined by $\Frob{}(n) = -n$.
By contract, the etale group scheme $\ZZ \times \ZZ$ with trivial Galois action
 does admit invisible quasicharacter sheaves;
 in particular, there are non-trivial quasicharacter sheaves on this group
with trivial trace of Frobenius.\footnote{Move this example: Define $g: \ZZ\times \ZZ \to \EEx$ by $g((n_1,m_1), (n_2, m_2)) = \ell^{m_1-m_2 + n_1-n_2}$.
Then use the tehniques of Section~\ref{sec:etale} to build a non-trivial character sheaf from the pair $(f=1,g)$ with trivial trace of Frobenius.}

@@

We also want to situate our terminology in a historical context.
It is not uncommon to refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}.
As explained above, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
We note that there are no invisble quasicharacter sheaves on such groups.

@@

%\begin{acknowledgements}
\subsection*{Acknowledgements}
We thank Masoud Kamgarpour, Pramod Achar, and Hadi Salmasian
for allowing us to hijack much of our Research in Teams meeting at BIRS into a discussion of
quasicharacter sheaves; their knowledge and help have been invaluable.
We also thank Takashi Suzuki, who made some enormously helpful observations and suggestions.
%
Finally, we gratefully acknowledge the financial support of the Pacific Institute for the Mathematical Sciences
and the National Science and Engineering Research Council,
as well the hospitality of the Banff International Research Station during our weeklong stay in May 2012.

@@

\begin{definition}\label{def:trFrob}
The function
\begin{align*}
\TrFrob{G}: \QCiso{G} &\to G(\Fq)^* \\
\qcs{L} &\mapsto t_{\qcs{L}}
\end{align*}
is called \emph{Trace of Frobenius}.\todo{Does this really warrant a declared definition?
Instead, we can refer to this very short section, when needed.}
\end{definition}

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the group homomorphism
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $s_\mathcal{L} \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $s_\mathcal{L} = s_\mathcal{L'}$ when $\qcs{L} \iso \qcs{L'}$, and that $s_{\qcs{L}\otimes \qcs{L'}} = s_{\qcs{L}} \, s_\qcs{L'}$
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.
We claim that $s_\qcs{L}$ is fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Observe:
\begin{equation}\label{ftot}
f\vert_{G(\Fq)} = t_\qcs{L},
\end{equation}
which is independent of the choice of $\{ v_x \in \gfqcs{L}_x \tq x\in G(\bFq)\}$.
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
\begin{align*}
S_G: \QCiso{G} &\to \Hh^0(\Weil{},\Hh^2(G(\bFq),\EEx))\\
\qcs{L} &\mapsto s_{\qcs{L}}.
\end{align*}

\begin{definition}\label{def:invisible}
Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We say $\qcs{L}$ is \emph{visible} if $s_\qcs{L} \in H^0(\Weil{},H^2(G(\bFq),\EEx))$ is trivial; otherwise, we say $\qcs{L}$ is \emph{invisible}.
\end{definition}

This terminology will be justified later when\todo{Point somewhere.} we show that a non-trivial quasicharacter sheaf $\qcs{L}$ on $G$ is invisible if and only if $\trFrob{\qcs{L}}$ is trivial.

@@

\begin{proposition} \label{prop:etale-iso}
	If $G$ is an \'etale commutative group scheme over $\Fq$
  	then
  	\[
  	\QCiso{G} \iso \Hh^2(G(\bFq)\rtimes \Weil{}, \EEx),
  	\]
  	functorially,
  	in the sense of Proposition~\ref{prop:functorialG}.
\end{proposition}

The proof of Proposition~\ref{prop:etale-iso} will require a few lemmas,
which we assemble now.\todo{This section is undergoing massive reconstruction. Don't read.}
%
%We begin with a simple result relating duals, invariants and coinvariants.

%Now we see that the Trace of Frobenius is surjective.

\begin{lemma}\label{lem:etale-trace}
If $G$ is an \'etale commutative group scheme over $\Fq$
then Trace of Frobenius
\[
\TrFrob{G} : \QCiso{G} \to G(\Fq)^*
\]
is a surjective group homomorphism.
\end{lemma}

@@

%Likewise, $S_G$ is surjective.

\begin{lemma}\label{lem:etale-2cocycle}
    If $G$ is an \'etale commutative group scheme over $\Fq$
    then the group homomorphism
    \[
    S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)),
    \]
    defined in Section~\ref{sec:invisible}, is surjective.
\end{lemma}

\begin{proof}
Pick $[g]\in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$. Then $g\in Z^2(G(\bFq),\EEx)$ is fixed by Frobenius, so $\,^{\Frob{}}g = g h$ for some $h\in B^2(G(\bFq),\EEx)$. Accordingly, $h(x,y) = \frac{f(x+y)}{f(x) f(y)}$ for some $f : G(\bFq) \to \EEx$, so
\begin{equation}\label{gf}
\frac{g(\Frob{}(x),\Frob{}(y))}{g(x,y)} = \frac{f(x+y)}{f(x) f(y)}
\end{equation}
for all $x,y\in G(\bFq)$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$ and define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x \otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = g(x,y) 1 \otimes 1$.
Since $g\in Z^2(G(\bFq),\EEx)$, the indexed set of isomorphisms $\{ \mu_{x,y} \tq x,y \in G(\bFq) \}$ satisfies \ref{qc.2}.
Define $\phi_x : \gqcs{L}_{\Frob{}(x)}\to \gqcs{L}_x$ by $\phi_x(1) = f(x)$.
Then set of isomorphisms $\{ \phi_{x} \tq x \in G(\bFq) \}$ satisfies \ref{qc.3} by \eqref{gf}. In this way, the cocycle $g\in Z^0(\Weil{}, Z^2(G(\bFq),\EEx))$ defines a quasicharacter sheaf $\qcs{L}$. It is clear that $\qcs{L}$ maps to $[g] \in \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))$ under \eqref{2cocycle}.
\end{proof}

\begin{lemma}\label{lem:trivial}
\[
\ker T_{G} \cap \ker S_{G} = 0
\]
\end{lemma}

\begin{proof}
Suppose $\trFrob{\qcs{L}} =1$ and $s_\qcs{L} =1$
Let $f$ and $g$ be as above.
Since the class of $g$ is trivial, $g(x,y) = \frac{h(x+y)}{h(x) h(y)}$ for some $h : G(\bFq) \to \EEx$;
since the class of $f$ is trivial, $f(x) = \frac{k(\Frob{G}(x))}{k(x)}$ for some $k : G(\bFq) \to \EEx$.
\todo{Actually, the lemma is false.}
\end{proof}

Now we can prove Proposition~\ref{prop:etale-iso}.

\begin{proof}[Proof of Proposition~\ref{prop:etale-iso}]
By Lemma~\ref{lem:dual-inv}, $G(\Fq)^* \iso \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))$. Composing this isomorphism with the Trace of Frobenius gives
\[
T_{G}: \QCiso{G}\to  \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx)),
\]
which is surjective, Lemma~\ref{lem:etale-trace}.
Combining this with
\[
S_{G} : \QCiso{G} \to \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx))
\]
gives
\[
\QCiso{G} \mathop{\longrightarrow}\limits^{R_G} \Hh^1(\Weil{},\Hh^1(G(\bFq),\EEx))\oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
The Hochschild-Serre spectral sequence provides
\[
\Hh^2(G(\bFq)\rtimes \Weil{},\EEx)  \iso \Hh^1(\Weil{}, \Hh^1(G(\bFq),\EEx)) \oplus \Hh^0(\Weil{}, \Hh^2(G(\bFq),\EEx)).
\]
Composing these last two maps defines
\begin{equation}
\QCiso{G}\to \Hh^2(G(\bFq)\rtimes \Weil{},\EEx).
\end{equation}
To see that this homomorphism is an isomorphism, observe that $S_G$ and $T_G$ are surjective and $\ker S_G \cap \ker T_G = 0$.
\end{proof}

@@

Consider the short exact sequence in the category of $\QC$-schemes
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is also a $\QC$-scheme,
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
%0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{G^0}$
and $\QCiso{\pi_0(G)}$
in Sections~\ref{sec:connected} and \ref{sec:etale}.
In this section we show that \eqref{eq:pi2} is exact.\todo{I wonder if $(F_1,F_2)$, see \ref{prop:functorialG}, form an adjoint pair.}

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth, commutative group scheme over
  $\Fq$ then Trace of Frobenius
  \[
  \TrFrob{G} : \QCiso{G} \to G(\Fq)^*
  \]
  is a functorial surjection with kernel $\Hh^0(\Weil{}, \Hh^2(\pi_0({\bar G}), \EE))$ and a canonical section.\todo{I wonder if $(F_1,F_2)$, see \ref{prop:functorialG}, form an adjoint pair. That seems to be what this theorem is saying.}
\end{theorem}

@@

\end{document}
%sagemathcloud={"zoom_width":125}
